market-justice-preferences

### 2024-07-04 02-antecedentes.qmd

The justification of social inequality based on market-type criteria has been conceptualized as the individuals' adherence to the deservingness of social goods and services (such as health, education, and pensions) based on prices and individuals' ability to pay [@boltanski_new_2005; @lane_market_1986; @streeck_citizens_2012]. Research on social stratification beliefs, which explore individual perceptions of who deserves what and why [@kluegel_beliefs_1987], highlights that people's explanations and justifications of social inequality are closely tied to their judgments of deservingness. The influence of ideologies [@wegener_dominant_1995] and cultural schemas [@homan_being_2017] is pivotal in shaping these explanations by offering symbolic representations that frame societal structures and expectations. While significant attention has been paid to wage inequality, income distribution, and payment differentials in the literature [@castillo_legitimacy_2011; Evans et al., 2010; @jasso_how_1999 ; @shariff_income_2016], there has been less examination of public beliefs about which life domains should be governed by market relations [@lindh_bringing_2023] and even less about children's acceptance or rejection of these market principles. This oversight is notable given the extensive encroachment of market logic into public goods, welfare policy, and social services over the past five decades [@centeno_arc_2012; @harvey_breve_2015], affecting areas such as pensions, health services, and education.

@@

Meritocratic beliefs can cover two types of subjective processes: preferences and perceptions [@castillo_multidimensional_2023]. While meritocratic preferences refer to a justification of distribution based on merit criteria (effort and talent), the perception of meritocracy refers to how individuals view and understand the concept of meritocracy in their society [@duru-bellat_who_2012; @castillo_meritocracia_2019]. The perception can vary greatly depending on individual experiences, as well as social, economic, and cultural background. Some people may see meritocracy as a fair and just system that allows anyone to succeed based on their abilities and hard work. In contrast, others may view it as a myth or a cover for existing power dynamics and inequality, serving to maintain and even reinforce inequality [@lampert_meritocratic_2013; @mijs_paradox_2019]. Some studies have analyzed how those with greater privileges believe more in meritocracy (Reynolds & Chan 2014), how greater economic inequality increases meritocratic beliefs [@mijs_paradox_2019], and how larger inequality decreases it [@morris_representing_2022].

>>>>

The justification of social inequality based on market-type criteria has been conceptualized as the individuals' adherence to the deservingness of social goods and services (such as health, education, and pensions) based on prices and individuals' ability to pay [@boltanski_new_2005; @lane_market_1986; @streeck_citizens_2012]. Research on social stratification beliefs, which explore individual perceptions of who deserves what and why [@kluegel_beliefs_1987], highlights that people's explanations and justifications of social inequality are closely tied to their judgments of deservingness. The influence of ideologies [@wegener_dominant_1995] and cultural schemas [@homan_being_2017] is pivotal in shaping these explanations by offering symbolic representations that frame societal structures and expectations. While significant attention has been paid to wage inequality, income distribution, and payment differentials in the literature [@castillo_legitimacy_2011; @kelley_legitimate_2021; @jasso_how_1999 ; @shariff_income_2016], there has been less examination of public beliefs about which life domains should be governed by market relations [@lindh_bringing_2023] and even less about children's acceptance or rejection of these market principles. This oversight is notable given the extensive encroachment of market logic into public goods, welfare policy, and social services over the past five decades [@centeno_arc_2012; @harvey_breve_2015], affecting areas such as pensions, health services, and education.

@@

Meritocratic beliefs can cover two types of subjective processes: preferences and perceptions [@castillo_multidimensional_2023]. While meritocratic preferences refer to a justification of distribution based on merit criteria (effort and talent), the perception of meritocracy refers to how individuals view and understand the concept of meritocracy in their society [@duru-bellat_who_2012; @castillo_meritocracia_2019]. The perception can vary greatly depending on individual experiences, as well as social, economic, and cultural background. Some people may see meritocracy as a fair and just system that allows anyone to succeed based on their abilities and hard work. In contrast, others may view it as a myth or a cover for existing power dynamics and inequality, serving to maintain and even reinforce inequality [@lampert_meritocratic_2013; @mijs_paradox_2019]. Some studies have analyzed how those with greater privileges believe more in meritocracy [@reynolds_perceptions_2014], how greater economic inequality increases meritocratic beliefs [@mijs_paradox_2019], and how larger inequality decreases it [@morris_representing_2022].

### 2024-07-04 06-conclusion.qmd

Our research aimed to examine the relationship between the perception of meritocracy and the justification of market justice preferences among eighth-grade students in Chile. From our knowledge this is the first study addressing market justice preferences at school level, finding that students exhibit large preferences for market justice (about one third agree or strongly agree with them), a share that is quite larger when compared with evidence in adult population. Perception of meritocracy is also high, particularly when it comes to the reward of effort, with an striking eighty six percent of agreement. Regarding these descriptive indicators, a a first message coming from this study is that there is something particular in school-age when it comes to perceptions and preferences about inequality. This can relate to development aspects (Kim & Gewirtz, 2019, Rizzo & Killen, 2020) but it could be something specific to the Chilean case, which requires further comparative as well as longitudinal research in the area. In any case, teaching about the origins and characteristics of social inequality is not part of the chilean school curriculum, calling for a more dedicated consideration of these topics in areas such as citizensip education as well as in history and social science education.

@@

The consideration of the school context opened some caveats in this research. We found contrary evidence for the association between status, school performance, and market justice preferences: high status and high achievement at the school level are associated with less market justice preferences. This is opposed to the rational-interest argument that suggests that those doing better - both economically and academically- would demand less redistribution (Gonthier,2016; Runst, 2018; Bullock, 2021) and therefore are expected to challenge market mechanisms. A possible explanation is that education contributes to forming a critical view of markets in redistribution, nevertheless, further research in this area is needed to assert the specific mechanisms that might be playing a role. All in all, from this study we know that school contexts are not trivial in this regard, calling for larger attention to school-age research on the formation of inequality preferences and beliefs.

>>>>

Our research aimed to examine the relationship between the perception of meritocracy and the justification of market justice preferences among eighth-grade students in Chile. From our knowledge this is the first study addressing market justice preferences at school level, finding that students exhibit large preferences for market justice (about one third agree or strongly agree with them), a share that is quite larger when compared with evidence in adult population. Perception of meritocracy is also high, particularly when it comes to the reward of effort, with an striking eighty six percent of agreement. Regarding these descriptive indicators, a a first message coming from this study is that there is something particular in school-age when it comes to perceptions and preferences about inequality. This can relate to development aspects [@kim_socioeconomic_2020; @rizzo_children_2020] but it could be something specific to the Chilean case, which requires further comparative as well as longitudinal research in the area. In any case, teaching about the origins and characteristics of social inequality is not part of the chilean school curriculum, calling for a more dedicated consideration of these topics in areas such as citizenship education as well as in history and social science education.

@@

The consideration of the school context opened some caveats in this research. We found contrary evidence for the association between status, school performance, and market justice preferences: high status and high achievement at the school level are associated with less market justice preferences. This is opposed to the rational-interest argument that suggests that those doing better - both economically and academically- would demand less redistribution [@gonthier_parallel_2017; @runst_does_2018; @bullock_education_2021] and therefore are expected to challenge market mechanisms. A possible explanation is that education contributes to forming a critical view of markets in redistribution, nevertheless, further research in this area is needed to assert the specific mechanisms that might be playing a role. All in all, from this study we know that school contexts are not trivial in this regard, calling for larger attention to school-age research on the formation of inequality preferences and beliefs.
